Fractional Types

### 2019-10-31 fracGC/p.tex
more on intro

\paragraph*{Quantum and Reversible Computing.} The Toffoli gate is
known to be universal for classical reversible
circuits~\cite{Toffoli:1980}. Adding just one gate (the Hadamard gate)
produces a universal set of primitives for quantum
circuits~\cite{hadtoffuniv}. The ``only'' difference between the two
circuit models is that quantum circuits can process superpositions of
values (waves) in one step whereas classical circuits lack this form
of parallelism. Most importantly, structurally the two circuits models
are identical and one can derive properties valid for both families by
focusing on the simpler classical model.

Instead of using the Toffoli gate as a universal primitive for
reversible classical circuits, one can exploit the full force of type
theory and category theory by expressing reversible classical
computations as \emph{isomorphisms over finite
  types}~\cite{Fiore:2004,James:2012:IE:2103656.2103667} or
\emph{equivalences over
  groupoids}~\cite{DBLP:conf/esop/CaretteS16}. This perspective
exposes interesting mathematical structure in reversible computations
that we will exploit to solve the ``ancilla problem'' explained next.

\paragraph*{Temporary Storage using Ancilla Bits.}

>>>>

\paragraph*{Quantum and Reversible Computing.} The Toffoli gate is
known to be universal for classical reversible
circuits~\cite{Toffoli:1980}. Adding just one gate (the Hadamard gate)
produces a universal set of primitives for quantum
circuits~\cite{hadtoffuniv}. The ``only'' difference between the two
circuit models is that quantum circuits can process superpositions of
values (waves) in one step whereas classical circuits lack this form
of parallelism. Most importantly, structurally the two circuits models
are identical and one can derive properties valid for both families by
focusing on the simpler classical model. In fact, classical reversible
computations are often used as ``subroutines'' of quantum
computations.

Instead of using the Toffoli gate as a universal primitive for
reversible classical circuits, one can leverage the full force of type
theory and category theory by expressing reversible classical
computations as \emph{isomorphisms over finite
  types}~\cite{Fiore:2004,James:2012:IE:2103656.2103667} or
\emph{equivalences over
  groupoids}~\cite{DBLP:conf/esop/CaretteS16}. This perspective
exposes interesting mathematical structure in reversible computations
that we will exploit to solve the ``ancilla problem'' explained next.

\paragraph*{Temporary Storage using Ancilla Bits.} The universality of
the Toffoli gate for classical reversible computing and of the
combination of the Toffoli and Hadamard gates for quantum computing
should not distract from efficiency and safety concerns. The theorems
proving universality (i) assume that an unlimited amount of temporary
storage may be
used~\cite{Toffoli:1980,DBLP:conf/innovations/AaronsonGS17}, and (ii)
that this temporary storage is returned to its initial state before
de-allocation. The latter condition is necessary in order maintain
symmetry between input and output in reversible circuits and to avoid
decoherence problems in quantum computing, Unfortunately, in current
proposals for reversible and quantum programming languages
(e.g. Quipper~\cite{Green:2013:QSQ:2491956.2462177}), this condition
is at the programmer's discretion.
